---
title: "2SFCA Lecture"
author: "Ivor Mardesic"
date: "12/10/2020"
output: html_document
---

```{r Markdown setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
options(scipen = 999)
rm(list=ls())
#https://github.com/adam-p/markdown-here/wiki/Markdown-Cheatsheet
```


# Spatio-temporal accessibility analysis using the 2SFCA methodology - Ivor Mardesic

In this Markdown document the 2SFCA methodology is implemented on the example of accessibility to hospitals for the citizens of the City of Zurich.


## Prepare workspace
We will use three R packages.

[Sf](https://cran.r-project.org/web/packages/sf/sf.pdf) for managing spatial data.
[tmap](https://cran.r-project.org/web/packages/tmap/tmap.pdf) for visualizations.
[cppRouting](https://cran.r-project.org/web/packages/cppRouting/cppRouting.pdf) for calculating isochrones for points of interest.
```{r R Packages}
# Required packages
# install.packages("sf")
# install.packages("tmap")
# install.packages("cppRouting")
library(sf)
library(tmap)
library(cppRouting)
library(sp)
library(tidyverse)
```

## Import data

[Zurich hospitals](https://www.stadt-zuerich.ch/geodaten/download/Spital?format=10007)

[Road network](https://maps.zh.ch/)

[Zurich Stadtkreise](https://www.stadt-zuerich.ch/geodaten/download/Stadtkreise?format=10007)

Demographic data - GIUZ server
```{r Data import, echo=F}
# Zurich kreise shapefile
kreis <- st_read(dsn = paste(getwd(), "/Data/kreise/data/stzh.adm_stadtkreise_a_polygon.shp", sep = "/"))
zurich <- st_union(kreis) # Merge kreise to derive city of Zurich polygon


# Zurich hospitals shapefile
services <- st_read(dsn = "C:/Users/Ivor/Documents/R/Projects/2SFCA/Data/spital/data")
services <- services[!duplicated(services$adresse),]
services <- services[c("adresse", "geometry")]
services$road_network_position <- services$geometry
services$foot_network_position <- services$geometry

# Swiss Bureau of Statistics 2013 Hectare population dataset
demographic_data <- read.csv(file = paste(getwd(),  "/Data/MSc/volkszaehlung_2013_statpop/rawdata/gd-b-00.03-13-vz2013statpopb/STATPOP2013B.csv", sep = "/"))
demographic_data <- demographic_data[, c("X_KOORD", "Y_KOORD", "B13BTOT", "RELI")] # subset useful things
demographic_data <- st_as_sf(x = demographic_data, coords = c("X_KOORD", "Y_KOORD")) # spatial points
st_crs(demographic_data) <- "EPSG:21781" # designate CRS
demographic_data <- st_transform(x = demographic_data, crs = st_crs(zurich)) # equalize CRS between data
names(demographic_data) <- c("poptotal", "PK", "geometry") # rename columns

## Subset demographic data to only cover the City of Zurich
population <- st_intersection(x = demographic_data, y = zurich) 
population <- population[-4294,]
population$road_network_position <- c(NA) # will be populated later
population$foot_network_position <- c(NA) # will be populated later
population$PK <- as.integer(1:nrow(population)) # our own PK list for the Zurich subset


# Road road_network dataset
road_network <- st_read(dsn = paste(getwd(), "/Data/MSc/Strassenentz/strasen/Strassennetz/TBA_STR_ACHS_L.shp", sep = "/"))
road_network <- st_transform(x = road_network, crs = st_crs(zurich)) # set crs
road_network <- road_network[, c("LENGTH", "geometry")] # subset usefull columns
road_network$time_cost <- as.integer(0) # input needed for isochrones
road_network$PK <- 1:nrow(road_network)

# Foot road_network dataset
foot_network <- st_read(dsn = paste(getwd(), "/Data/fussweg/Fussweg.shp", sep = "/"))
foot_network <- st_transform(x = foot_network, crs = st_crs(zurich)) # set crs
foot_network$PK <- 1:nrow(foot_network)
foot_network$time_cost <- as.integer(0) # input needed for isochrones
foot_network$incline <- as.numeric(0)
## Multilinestring to string
foot_network <- st_cast(x = foot_network, to = "LINESTRING")

# 
# tmap_mode("view")
# tm_shape(zurich)+
#   tm_polygons(col = "white")+
#   tm_shape(population)+
#   tm_symbols(col = "red", palette = "Pastel1", 
#              size = "poptotal", scale = 1, 
#              border.col = NA)+
#   tm_shape(services)+
#   tm_symbols(col = "blue")+
#   tm_shape(road_network)+
#   tm_lines(col = "black")
```

## Foot network incline

Calculate the network segment inclines.

Challenge! One link has a downhill and uphill version, code must recognize which one is used.
For later...

Incline formula; divide length by height difference.

First, calculate segment length (haversine, or cheat?).

```{r foot network inclines}
library(sp)

foot_network$delta_z <- abs(foot_network$HOEHE_ANFA - foot_network$HOEHE_ENDE)
foot_network$length <- st_length(foot_network) %>% as.numeric()

foot_network$incline <- foot_network$delta_z/foot_network$length
```


# Data pre-processing
Before we can implement the 2sfca, the data has to be of an appropriate format and depth. We need to prepare the road network dataset for calculating isochrones; catchment areas reachable from a location within a given time budget and network speeds.

The implementation of isochrones will be done using the cppRouting package which integrates the theories of Edger Dijkstra, who worked with and developed graph theory applications in computer science. The Dijkstra algorithm is used for routing between points on a network and is well established in transport sciences; and subsequently, very efficient. The cppRouting package integrates it with R and with some workarounds, we can implement it with spatial data.

Before we can develop isochrones we cannot do the 2SFCA methodology.

Furthermore, for proper isochronal analysis we need to know, for each service and demand point, what is the closest road and "snap" this point of interest to that road. 

## Road network vertices
The road network is a list of all roads with their geometry. We have to find where do the geometries interesct and declare these intersections as a vertex per the graph theory (a graph is a set of edges and vertices). We have to save in the road network the start and end vertex of a road (edge) as the index of the vertices in a separate spatial data frame. The vertices dataframe will contain the PK (the index) and the X and Y coordinate of the vertex.

In our dataset, every start and end coordinate of a line is a vertex. We need to extract these, assign them a unique ID, and record which edges share them to remove duplicates.

```{r Network vertices, warning=F}


road_network$start_vertex <- st_line_sample(road_network, sample = 0) # st_line_sample with sample = 0 provides the first point of a linestring
road_network$start_vertex <- st_cast(x = road_network$start_vertex, "POINT") # define proper spatial object class; much better to be exact than wonder why is the code not working down the line
road_network$end_vertex <- st_line_sample(road_network, sample = 1) # st_line_sample with sample = 1 provides the last point of a linestring
road_network$end_vertex <- st_cast(x = road_network$end_vertex, "POINT") # define proper spatial object class
#road_network$intersects <- c(NA)

x <- c(st_coordinates(road_network$end_vertex)[,1], st_coordinates(road_network$start_vertex)[,1]) 
y <- c(st_coordinates(road_network$end_vertex)[,2], st_coordinates(road_network$start_vertex)[,2])

road_vertices <- data.frame("x" = x, # get X coord of all road_vertices
                       "y" = y, # get y coord of all road_vertices
                       stringsAsFactors = F)

road_vertices <- st_as_sf(x = road_vertices, coords = c("x", "y"), remove = F) # turn into spatial points df
st_crs(road_vertices) <- st_crs(road_network) # regulate crs! ALWAYS TAKE NOTE OF CRS

# remove duplicates
road_vertices <- road_vertices[!duplicated(road_vertices$geometry),]
rm(x,y)

road_vertices$PK <- 1:nrow(road_vertices)
row.names(road_vertices) <- road_vertices$PK # re-align PK and row numbers in View()




# Relate each road with a vertex
road_network$from <- c(NA)
road_network$to <- c(NA)

road_network$from <- st_equals(x = road_network$start_vertex, y = road_vertices$geometry) # which vertex starts the road?
road_network$to <- st_equals(x = road_network$end_vertex, y = road_vertices$geometry) # which vertex ends the road?


#### visual inspection####
# tmap_mode("view")
# 
# tm_shape(road_network$geometry)+
#   tm_lines(col = "black", lwd = 1)+
#   tm_basemap(NULL)+
#   tm_shape(road_vertices$geometry)+
#   tm_symbols(size = 0.001, col = "red")


```
## Foot network vertices
Hodas data

```{r foot_network vertices, warning=F}


foot_network$start_vertex <- st_line_sample(foot_network, sample = 0) # st_line_sample with sample = 0 provides the first point of a linestring
foot_network$start_vertex <- st_cast(x = foot_network$start_vertex, "POINT") # define proper spatial object class; much better to be exact than wonder why is the code not working down the line
foot_network$end_vertex <- st_line_sample(foot_network, sample = 1) # st_line_sample with sample = 1 provides the last point of a linestring
foot_network$end_vertex <- st_cast(x = foot_network$end_vertex, "POINT") # define proper spatial object class
#foot_network$intersects <- c(NA)

x <- c(st_coordinates(foot_network$end_vertex)[,1], st_coordinates(foot_network$start_vertex)[,1]) 
y <- c(st_coordinates(foot_network$end_vertex)[,2], st_coordinates(foot_network$start_vertex)[,2])

foot_vertices <- data.frame("x" = x, # get X coord of all vertices
                       "y" = y, # get y coord of all vertices
                       stringsAsFactors = F)

foot_vertices <- st_as_sf(x = foot_vertices, coords = c("x", "y"), remove = F) # turn into spatial points df
st_crs(foot_vertices) <- st_crs(foot_network) # regulate crs! ALWAYS TAKE NOTE OF CRS

# remove duplicates
foot_vertices <- foot_vertices[!duplicated(foot_vertices$geometry),]
rm(x,y)

foot_vertices$PK <- 1:nrow(foot_vertices)
row.names(foot_vertices) <- foot_vertices$PK # re-align PK and row numbers in View()




# Relate each road with a vertex
foot_network$from <- c(NA)
foot_network$to <- c(NA)

foot_network$from <- st_equals(x = foot_network$start_vertex, y = foot_vertices$geometry) # which vertex starts the road?
foot_network$to <- st_equals(x = foot_network$end_vertex, y = foot_vertices$geometry) # which vertex ends the road?


#### visual inspection####
tmap_mode("view")
# 
# tm_shape(foot_network$geometry)+
#   tm_lines(col = "black", lwd = 1)+
#   tm_basemap(NULL)+
#   tm_shape(foot_vertices$geometry)+
#   tm_symbols(size = 0.001, col = "red")


```

## Snap population and service points to road network
Now that we know the vertices of our transport network, we need to snap populations and hospitals to the nearest vertex. Alternatively (e.g. with network analyst in ARcGIS), the snapping can be done to the nearest edge. However, the cppRouting package uses vertices as the starting points for calculating isochrones.


```{r Network position snap, warning=F, echo=F}

# snap hospitals to roads
services$road_vertex_id <- st_nearest_feature(x = services, y = road_vertices)
services$road_network_position <- st_geometry(road_vertices)[services$road_vertex_id]


# snap demographic data to roads
population$road_vertex_id <- st_nearest_feature(x = population, y = road_vertices)
population$road_network_position <- st_geometry(road_vertices)[population$road_vertex_id]


# no snapping
#   tm_shape(population$geometry)+
#   tm_symbols(col = "red", palette = "Pastel1",
#              size = 0.01,
#              border.col = NA)+
#   tm_legend(show = F)+
#   tm_shape(services$geometry)+
#   tm_symbols(col = "blue")+
#   tm_shape(network$geometry)+
#   tm_lines(col = "black")
# 
# # with snapping
#   tm_shape(population$road_network_position)+
#   tm_symbols(col = "red", palette = "Pastel1",
#              size = 0.1,
#              border.col = NA)+
#   tm_legend(show = F)+
#   tm_shape(services$road_network_position)+
#   tm_symbols(col = "blue")+
#   tm_shape(network$geometry)+
#   tm_lines(col = "black")
```

##Snap foot network

```{r foot network position snap, warning=F, echo=F}

# snap hospitals to roads
services$foot_vertex_id <- st_nearest_feature(x = services, y = foot_vertices)
services$foot_network_position <- st_geometry(foot_vertices)[services$foot_vertex_id]


# snap demographic data to roads
population$foot_vertex_id <- st_nearest_feature(x = population, y = foot_vertices)
population$foot_network_position <- st_geometry(foot_vertices)[population$foot_vertex_id]



# # no snapping
#   tm_shape(population$geometry)+
#   tm_symbols(col = "red", palette = "Pastel1",
#              size = 0.01,
#              border.col = NA)+
#   tm_legend(show = F)+
#   tm_shape(services$geometry)+
#   tm_symbols(col = "blue")+
#   tm_shape(foot_network$geometry)+
#   tm_lines(col = "black")
# 
# # with snapping
#   tm_shape(population$foot_network_position)+
#   tm_symbols(col = "red", palette = "Pastel1",
#              size = 0.1,
#              border.col = NA)+
#   tm_legend(show = F)+
#   tm_shape(services$foot_network_position)+
#   tm_symbols(col = "blue")+
#   tm_shape(foot_network$geometry)+
#   tm_lines(col = "black")
```



# Derive Isochrones
Having prepared our data for implementing isochrones, all that is left to be done is assign cost factors to roads, define network speeds (e.g. 5km/h for walking speed), and build a "network dataset"; that is, a transport network graph in this particular implementation.

## Pre-isochrone input parameters
```{r Movement speed and network costs input}
# time budget and travel costs for the network
movement_speed <- 5 #in km/hour


meter_per_minute <- (movement_speed*1000)/60 # how many meters do we cross per minute with given movement speed in km/h
road_network$time_cost <- road_network$LENGTH/meter_per_minute # how many minutes does it cost to cover the edge
foot_network$time_cost <- foot_network$length/meter_per_minute # how many minutes does it cost to cover the edge

rm(meter_per_minute)
```

## Dijkstra graphs
Dijkstra algorithm graph. Data needs specific format to properly make graph.
https://github.com/vlarmet/cppRouting#package-presentation


```{r Dijkstra algorithm graph}
# initialize network graph object
dijkstra_roads <- road_network[,c("from", "to", "time_cost")]
dijkstra_footpaths <- foot_network[,c("from", "to", "time_cost")]

# define geometries (i.e. column)
st_geometry(dijkstra_roads) <- NULL
st_geometry(dijkstra_footpaths) <- NULL

# coordinates of all network vertices (needed for dijkstra)
road_coord <- road_vertices[,c("PK", "x", "y")]
st_geometry(road_coord) <- NULL
foot_coord <- foot_vertices[,c("PK", "x", "y")]
st_geometry(foot_coord) <- NULL

# dijkstra graphs
road_graph <- makegraph(dijkstra_roads, directed = F, coords = road_coord) # non-directed; else algorithm looks only one direction
foot_graph <- makegraph(dijkstra_footpaths, directed = F, coords = foot_coord) # foot network version
```


## Isochrone example
Isochrones are catchment areas around points of interest which encompass all space reachable through the network at a given time budget and movement speed.

Isochrones use network distances which is preferable to euclidean distances (e.g. buffers). Network distances better model spatial reality and actual reachability, and take into account natural barriers (e.g. a river).

```{r Isochrone example}

# road_iso <- get_isochrone(road_graph, from = "500", lim = 10, long = T) # run isochrone for vertex 1823; lim determines time budget; long = T to give dataframe
# foot_iso <- get_isochrone(foot_graph, from = "500", lim = 10, long = T) # lim = time budget
#
# road_extent <- road_vertices[road_iso$node,] # get all reachable vertices from starting vertex
# road_extent <- st_multipoint(x = cbind(st_coordinates(road_extent)[,1], st_coordinates(road_extent)[,2])) # turn into multipoint
# foot_extent <- foot_vertices[foot_iso$node,] # get all reachable foot_vertices from starting vertex
# foot_extent <- st_multipoint(x = cbind(st_coordinates(foot_extent)[,1], st_coordinates(foot_extent)[,2])) # turn into multipoint
#
# # roadnet catchment
# road_catchment <- st_convex_hull(road_extent) # get convex hull from the multipoint feature
# road_catchment <- st_geometry(road_catchment) # force back into sfc object
# st_crs(road_catchment) <- st_crs(road_vertices) # manage the crs
# # footnet version
# foot_catchment <- st_convex_hull(foot_extent) # get convex hull from the multipoint feature
# foot_catchment <- st_geometry(foot_catchment) # force back into sfc object
# st_crs(foot_catchment) <- st_crs(foot_vertices) # manage the crs



#### visual inspection####
# tmap_mode("view")
#
# tm_shape(road_catchment)+
#   tm_polygons(col = "red")+
#   tm_shape(road_network$geometry)+
#   tm_lines(col = "black", lwd = 1)+
#   tm_basemap(NULL)+
#   tm_shape(road_vertices$geometry)+
#   tm_symbols(size = 0.0001, col = "blue")+
#   tm_shape(population)+
#   tm_symbols(size = 0.001, col = "green")+
#   tm_shape(road_vertices$geometry[as.integer(unique(road_iso$origin))])+
#   tm_symbols(size = 1, col = "purple")
#
#
# #footnet version
# tm_shape(foot_catchment)+
#   tm_polygons(col = "red")+
#   tm_shape(foot_network$geometry)+
#   tm_lines(col = "black", lwd = 1)+
#   tm_basemap(NULL)+
#   tm_shape(foot_vertices$geometry)+
#   tm_symbols(size = 0.0001, col = "blue")+
#   tm_shape(population)+
#   tm_symbols(size = 0.001, col = "green")+
#   tm_shape(foot_vertices$geometry[as.integer(unique(foot_iso$origin))])+
#   tm_symbols(size = 1, col = "purple")

```

## Data frame for storing results
We will create a dataframe for storing the results of the 2SFCA analysis.
```{r Results dataframe}
road_results_2sfca_services <- data.frame("location_id" = c(NA), # row id of the location in its perspective dataframe
                                         "catchment_poly" = c(NA), # the geometry of the isochrone catchment
                                         "road_vertex_id" = c(NA), # the id of the snapped to vertex
                                         "time_budget" = c(NA), # the time budget of the isochrone analysis
                                         stringsAsFactors = F)[1:nrow(services),] # populate rows

road_results_2sfca_population <- data.frame("location_id" = c(NA), 
                                           "catchment_poly" = c(NA),
                                           "road_vertex_id" = c(NA),
                                           "time_budget" = c(NA),
                                           stringsAsFactors = F)[1:nrow(population),]


#footnet ver
foot_results_2sfca_services <- data.frame("location_id" = c(NA), # row id of the location in its perspective dataframe
                                         "catchment_poly" = c(NA), # the geometry of the isochrone catchment
                                         "foot_vertex_id" = c(NA), # the id of the snapped to vertex
                                         "time_budget" = c(NA), # the time budget of the isochrone analysis
                                         stringsAsFactors = F)[1:nrow(services),] # populate rows

foot_results_2sfca_population <- data.frame("location_id" = c(NA), 
                                           "catchment_poly" = c(NA),
                                           "foot_vertex_id" = c(NA),
                                           "time_budget" = c(NA),
                                           stringsAsFactors = F)[1:nrow(population),]
```


### Fac/pop isochrones roadnet
Now that we have the road position of every point of interest, we can create isochrones for that location.
```{r Isochrone analysis, echo = F}
time_budget <- 10 # in minutes

# hospital isochrones
for(i in 1:nrow(services)) {
  
  road_iso <- get_isochrone(road_graph, # run isochrone for hospitals based on road graph
                            from = services$road_vertex_id[i], # starting points
                            lim = time_budget, long = T)  # time budget
  
  # calculate the road_extent of the isochrone
  road_extent <- road_vertices[road_iso$node,] # isochrones input
  road_extent <- st_multipoint(x = cbind(st_coordinates(road_extent)[,1], st_coordinates(road_extent)[,2]))
  
  # derive a catchment area from isochrone road_extent
  road_catchment <- st_convex_hull(road_extent) # get convex hull from extentable road_vertices
  road_catchment <- st_geometry(road_catchment) # force back into sfc object
  st_crs(road_catchment) <- st_crs(road_vertices)
  
  # save results
  road_results_2sfca_services$location_id[i] <- i
  road_results_2sfca_services$road_vertex_id[i] <- services$road_vertex_id[i]
  road_results_2sfca_services$time_budget[i] <- time_budget
  road_results_2sfca_services$catchment_poly[i] <- road_catchment
}

# population isochrones
for(i in 1:nrow(population)) {
  
  road_iso <- get_isochrone(road_graph,
                            from = population$road_vertex_id[i],
                            lim =time_budget, long = T) # run isochrone for hospitals
  
  # calculate the road_extent of the isochrone
  road_extent <- road_vertices[road_iso$node,]
  road_extent <- st_multipoint(x = cbind(st_coordinates(road_extent)[,1], st_coordinates(road_extent)[,2]))
  
  # derive a catchment area from isochrone road_extent
  road_catchment <- st_convex_hull(road_extent) # get convex hull from extentable road_vertices
  road_catchment <- st_geometry(road_catchment) # force back into sfc object
  st_crs(road_catchment) <- st_crs(road_vertices)
  
  # save results
  road_results_2sfca_population$location_id[i] <- i
  road_results_2sfca_population$road_vertex_id[i] <- population$road_vertex_id[i] # the id of the vertex assigned to the point; based on the road/foot vertices dataframe
  road_results_2sfca_population$time_budget[i] <- time_budget
  road_results_2sfca_population$catchment_poly[i] <- road_catchment
  
  if(i %in% seq(0,50000,by = 500)){print(i)}
  
}
```
### Fac/pop isochrones footnet


```{r footnet isochrones, echo = F}
time_budget <- 10 # in minutes

# hospital isochrones
for(i in 1:nrow(services)) {
  
  #graph part
  foot_iso <- get_isochrone(foot_graph, 
                            from = services$foot_vertex_id[i],
                            lim = time_budget, long = T) # run isochrone for hospitals
  
  # calculate the extent_foot of the isochrone
  foot_extent <- foot_vertices[foot_iso$node,]
  foot_extent <- st_multipoint(x = cbind(st_coordinates(foot_extent)[,1], st_coordinates(foot_extent)[,2]))
  
  # derive a foot_catchment area from isochrone extent
  foot_catchment <- st_convex_hull(foot_extent) # get convex hull from extentable vertices
  foot_catchment <- st_geometry(foot_catchment) # force back into sfc object
  st_crs(foot_catchment) <- st_crs(foot_vertices)
  
  # save results
  foot_results_2sfca_services$location_id[i] <- i
  foot_results_2sfca_services$foot_vertex_id[i] <- services$foot_vertex_id[i]
  foot_results_2sfca_services$time_budget[i] <- time_budget
  foot_results_2sfca_services$catchment_poly[i] <- foot_catchment
}

# population isochrones

for(i in 1:nrow(population)) {#nrow(population)
  
  foot_iso <- get_isochrone(foot_graph,
                            from = population$foot_vertex_id[i],
                            lim =time_budget, long = T) # run isochrone for hospitals
  
  # calculate the extent_foot of the isochrone
  foot_extent <- foot_vertices[foot_iso$node,]
  foot_extent <- st_multipoint(x = cbind(st_coordinates(foot_extent)[,1], st_coordinates(foot_extent)[,2]))
  
  # derive a foot_catchment area from isochrone extent
  foot_catchment <- st_convex_hull(foot_extent) # get convex hull from extentable foot_vertices
  foot_catchment <- st_geometry(foot_catchment) # force back into sfc object
  st_crs(foot_catchment) <- st_crs(foot_vertices)
  
  # save results
  foot_results_2sfca_population$location_id[i] <- i
  foot_results_2sfca_population$foot_vertex_id[i] <- population$foot_vertex_id[i]
  foot_results_2sfca_population$time_budget[i] <- time_budget
  foot_results_2sfca_population$catchment_poly[i] <- foot_catchment
  
  if(i %in% seq(0,50000,by = 500)){print(i)}
  
}
```

# 2SFCA analysis

Finally, we arrive to the reason why we are all here. 2SFCA implementation.

The preparatory steps are indespensible however, no matter the platform for implementation.

## 2SFCA - Step 1: Healthcare supply and demand at hospitals
We will calculate the supply-demand ratio at every hospital by dividing the a "supply capacity" with the total serviced population.

We will assume a per-doctor capacity of 24 patients in a day; and assign each hospital the arbitrary total of a 100 doctors. This data should ofcourse be informed and depending on your research can (and should) go into much detail; e.g. accessibility to dialysis, oncologists, respirators (Covid-19), etc. 

Therefore, the daily healthcare supply is 2400 people per hospital.


```{r 2SFCA Step 1}
people_per_doctor <- 24
doctors_per_hospital <- 100
capacity <- people_per_doctor*doctors_per_hospital # healthcare supply at a hospital

services$road_supply_demand_ratio <- as.numeric(0)
services$foot_supply_demand_ratio <- as.numeric(0)

road_results_2sfca_services$population_serviced <- c(NA) # record total population serviced by a hospital
road_results_2sfca_services$capacity <- capacity
road_results_2sfca_services$supply_demand_ratio <- c(NA) # supply-demand ratio

foot_results_2sfca_services$population_serviced <- c(NA) # record total population serviced by a hospital
foot_results_2sfca_services$capacity <- capacity
foot_results_2sfca_services$supply_demand_ratio <- c(NA) # supply-demand ratio

# road catchment analysis
for (i in 1:nrow(services)) {
  
  
  catchment <- road_results_2sfca_services$catchment_poly[[i]] # retrieve the geometry of the isochrone
  
  # supply-demand ratio calculation
  population_serviced <- st_intersects(population$geometry, catchment, sparse = F) #total pop in hospital catchment
  population_serviced <- sum(population$poptotal[population_serviced])
  supply_demand_ratio <- capacity/population_serviced
  
  # save results
  road_results_2sfca_services$supply_demand_ratio[i] <- supply_demand_ratio
  road_results_2sfca_services$population_serviced[i] <- population_serviced
  services$road_supply_demand_ratio[i] <- road_results_2sfca_services$supply_demand_ratio[i]
  
  
}

# foot catchment analysis
for (i in 1:nrow(services)) {
  
  
  catchment <- foot_results_2sfca_services$catchment_poly[[i]] # retrieve the geometry of the isochrone
  
  # supply-demand ratio calculation
  population_serviced <- st_intersects(population$geometry, catchment, sparse = F) #total pop in hospital catchment
  population_serviced <- sum(population$poptotal[population_serviced])
  supply_demand_ratio <- capacity/population_serviced
  
  # save results
  foot_results_2sfca_services$supply_demand_ratio[i] <- supply_demand_ratio
  foot_results_2sfca_services$population_serviced[i] <- population_serviced
  services$foot_supply_demand_ratio[i] <- foot_results_2sfca_services$supply_demand_ratio[i]
  
  
}
```

## 2SFCA - Step 2: Population accessibility to healthcare

Having calculated the supply-demand ratios at hospitals, we will find out what is the 2SFCA spatio-temporal accessibility score at each population point.

This is the second step of the 2fa and includes finding out which hospitals are reachable from every population point, and then summing up the supply-demand ratios of reachable hospitals.

What we get is an accessibility score which represents how much healthcare capacity is accessible to a person living in a certain population point. Some population points might have no hospitals reachable in the given time budget! The 2SFCA score will be 0 in that case.



```{r 2SFCA Step 2}
road_results_2sfca_population$accessibility_score <- c(NA) # 2sfca spatio-temporal accessibility score
road_results_2sfca_population$reachable_hospitals <- c(NA) # index of reachable hospitals
road_results_2sfca_population$poptotal <- c(NA) # total population at population point

foot_results_2sfca_population$accessibility_score <- c(NA) # 2sfca spatio-temporal accessibility score
foot_results_2sfca_population$reachable_hospitals <- c(NA) # index of reachable hospitals
foot_results_2sfca_population$poptotal <- c(NA) # total population at population point

# road net
for (i in 1:nrow(population)) {
  
  catchment <- road_results_2sfca_population$catchment_poly[[i]] # retrieve the geometry of the isochrone
  
  # reachable hospitals
  reachable_hospitals <- st_intersects(services$geometry, catchment, sparse = F) #total pop in hospital catchment
  road_results_2sfca_population$reachable_hospitals[i] <- list(which(reachable_hospitals==T)) # record the index of the hospitals
  
  # accessibility_score - sum of supply-demand ratios at reachable hospitals
  accessibility_score <- sum(road_results_2sfca_services[which(reachable_hospitals==T), "supply_demand_ratio"])
  
  # save results
  road_results_2sfca_population$accessibility_score[i] <- accessibility_score
  road_results_2sfca_population$poptotal[i] <- population$poptotal[i]

  
    if(i %in% seq(0,5000,by = 500)){print(i)} # just to keep track of progress
}


# foot net
for (i in 1:nrow(population)) {
  
  catchment <- foot_results_2sfca_population$catchment_poly[[i]] # retrieve the geometry of the isochrone
  
  # reachable hospitals
  reachable_hospitals <- st_intersects(services$geometry, catchment, sparse = F) #total pop in hospital catchment
  foot_results_2sfca_population$reachable_hospitals[i] <- list(which(reachable_hospitals==T)) # record the index of the hospitals
  
  # accessibility_score - sum of supply-demand ratios at reachable hospitals
  accessibility_score <- sum(foot_results_2sfca_services[which(reachable_hospitals==T), "supply_demand_ratio"])
  
  # save results
  foot_results_2sfca_population$accessibility_score[i] <- accessibility_score
  foot_results_2sfca_population$poptotal[i] <- population$poptotal[i]

  
    if(i %in% seq(0,5000,by = 500)){print(i)} # just to keep track of progress
}

```


# Results interpretation
We successfully implemented the 2SFCA methodology in R and derived accessibility scores for each population point.

WE saved these scores in a dataframe which can be related to the original population dataset through primary keys. What we are interested in now is the spatial variation in accessibility scores, as well as overall scores for the city of Zurich.

```{r Overall 2sfca scores}
print("2SFCA scores at population points foot")
summary(foot_results_2sfca_population$accessibility_score)
print("Supply-demand ratios at hospitals foot")
summary(foot_results_2sfca_services$supply_demand_ratio)
print("2SFCA scores at population points road")
summary(road_results_2sfca_population$accessibility_score)
print("Supply-demand ratios at hospitals road")
summary(road_results_2sfca_population$supply_demand_ratio)

population$foot_accessibility_score <- foot_results_2sfca_population$accessibility_score
population$road_accessibility_score <- road_results_2sfca_population$accessibility_score
```



